Inferring community characteristics in labelled networks

### 2021-05-28 nips/abstract.tex

\begin{abstract}
	Labelled networks form a very common and important class of data,
	naturally appearing in numerous applications in science and engineering.
	A typical inference goal is to determine how the vertex labels
	(or {\em features}) affect the network's graph structure. A standard
	approach has been to partition the network into blocks grouped
	by distinct values of the feature of interest. A block-based random
	graph model -- typically a variant of the stochastic block model --
	is then used to test for evidence of asymmetric behaviour within these
	feature-based communities. Nevertheless, the resulting communities
	often do not produce a natural partition of the graph. In this work
	we introduce a new generative model, the feature-first block model (FFBM),
	which is more effective at describing vertex-labelled undirected
	graphs and also facilitates the use of richer queries on labelled networks.
	We develop a Bayesian framework for inference with this model,
	and we present a method to efficiently sample from the posterior
	distribution of the FFBM parameters. The FFBM's structure is kept
	deliberately simple to retain easy interpretability of the parameter
	values. We apply the proposed methods to a variety of network data
	to extract the most important features along which the vertices
	are partitioned. The main advantages of the proposed approach are
	that the whole feature-space is used automatically, and features
	can be rank-ordered implicitly according to impact. Any features
	that do not significantly impact the high-level structure can be
	discarded to reduce the problem dimension. In cases where the vertex
	features available do not readily explain the community structure
	in the resulting network, the approach detects this and is protected
	against over-fitting. Results on several real-world datasets
	illustrate the performance of the proposed methods. 
\end{abstract}

>>>>

\begin{abstract}
	Labelled networks form a very common and important class of data,
	naturally appearing in numerous applications in science and engineering.
	A typical inference goal is to determine how the vertex labels
	(or {\em features}) affect the network's graph structure. A standard
	approach has been to partition the network into blocks grouped
	by distinct values of the feature of interest. A block-based random
	graph model -- typically a variant of the stochastic block model --
	is then used to test for evidence of asymmetric behaviour within these
	feature-based communities. Nevertheless, the resulting communities
	often do not produce a natural partition of the graph. In this work,
	we introduce a new generative model, the feature-first block model (FFBM),
	which is more effective at describing vertex-labelled undirected
	graphs and also facilitates the use of richer queries on labelled networks.
	We develop a Bayesian framework for inference with this model,
	and we present a method to efficiently sample from the posterior
	distribution of the FFBM parameters. The FFBM's structure is kept
	deliberately simple to retain easy interpretability of the parameter
	values. We apply the proposed methods to a variety of network data
	to extract the most important features along which the vertices
	are partitioned. The main advantages of the proposed approach are
	that the whole feature-space is used automatically, and features
	can be rank-ordered implicitly according to impact. Any features
	that do not significantly impact the high-level structure can be
	discarded to reduce the problem dimension. In cases where the vertex
	features available do not readily explain the community structure
	in the resulting network, the approach detects this and is protected
	against over-fitting. Results on several real-world datasets
	illustrate the performance of the proposed methods. 
\end{abstract}

### 2021-05-28 nips/ffbm.tex

Let $N$ denote the number of nodes, $B$ the number of blocks in the graph,
and ${\cal X}$ the set of all possible features.
We define the vector $x_i \in \Xcal^D$ as the feature vector of vertex $i$, 
where $D$ is the number of features associated with each vertex.
For the datasets we analyse, we deal with binary feature flags so $\Xcal = \{0, 1\}$. We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.

For the FFBM, we start with the feature matrix $X$ and generate a random
vector of block memberships $b \in [B]^N$. For each node $i$, the
block membership $b_i\in[B]$ is generated based on the feature
vector $x_i$, independently between nodes. The conditional
distribution of $b_i$ given $x_i$ also depends on a collection
of weight vectors $\theta=\{w_k\}_{k=1}^B$, where each
$w_k$ has dimension $D$. Specifically, 
the distribution of $b$ given $X$ and $\theta$ is,
%
\begin{equation}
	p(b| X, \theta) = \prod_{i=1}^{N} p(b_i | x_i, \theta) = \prod_{i=1}^{N} \phi_{b_i} (x_i; \theta)
	= \prod_{i=1}^{N} \frac{\exp\left(w_{b_i}^T x_i\right)}{\sum_{k=1}^{B} \exp \left( w_k^T x_i\right)}
\end{equation}
%
This is the form of a softmax; we deliberately exclude a bias term to ensure that the relationships we model are based on features and not information about the size of each detected block. 
We will later also find it convenient
to write the parameters $\theta$ as a $B \times D$ matrix of weights $W$.

@@

To complete the description of our Bayesian framework,
priors on $\theta$ and $\psi$ must also be specified. 
We place a Gaussian prior on $\theta$ with zero mean
and covariance matrix $\sigma^2_\theta I$, so that
each element of $\theta$ has an independent ${\cal N}(0,\sigma_\theta^2)$
prior, for a hyperparameter $\sigma_\theta^2$:
%
\begin{equation}
	p(\theta) \sim \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right).
	\label{eqn:theta-prior}
\end{equation}
%
Interestingly, this choice of prior leads to a particularly
simple distribution on the block membership vector $b$.
We show in Appendix~\ref{appdx:b|x} that,
after integrating out $\theta$,
$b$ is uniformly distributed:
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}.
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is an important simplification as evaluating $p(b | X)$ does not require an expensive Monte Carlo integration over $\theta$ nor does it require the exact value of $X$. 
\citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for 
the additional microcanonical SBM parameters $\psi$, which we adopt. 
The idea is to write the joint distribution on $(b, e, k)$ as a product of 
conditionals, $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. 
In our case, conditioning on $X$ is also necessary, leading to,
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b),
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
where we used the fact $\psi$ and $X$ are conditionally 
independent given $b$.  The exact form of the prior
$p(\psi|b)$ is described in Appendix \ref{appdx:prior}.
All that concerns the main argument is that it has
an easily computable form.

>>>>

Let $N$ denote the number of nodes, $B$ the number of blocks in the graph,
and ${\cal X}$ the set of values each feature can take.
We define the vector $x_i \in \Xcal^D$ as the feature vector of vertex $i$, 
where $D$ is the number of features associated with each vertex.
For the datasets we analyse, we deal with binary feature flags so $\Xcal = \{0, 1\}$. We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.

For the FFBM, we start with the feature matrix $X$ and generate a random
vector of block memberships $b \in [B]^N$. For each vertex $i$, the
block membership $b_i\in[B]$ is generated based on the feature
vector $x_i$, independently between vertices. The conditional
distribution of $b_i$ given $x_i$ also depends on a collection
of weight vectors $\theta=\{w_k\}_{k=1}^B$, where each
$w_k$ has dimension $D$. Specifically, 
the distribution of $b$ given $X$ and $\theta$ is,
%
\begin{equation}
	p(b| X, \theta) = \prod_{i=1}^{N} p(b_i | x_i, \theta) = \prod_{i=1}^{N} \phi_{b_i} (x_i; \theta)
	= \prod_{i=1}^{N} \frac{\exp\left(w_{b_i}^T x_i\right)}{\sum_{k=1}^{B} \exp \left( w_k^T x_i\right)}.
\end{equation}
%
$\phi_{b_i}$ is the form of a softmax activation function; we deliberately exclude a bias term to ensure that we only leverage feature information. 
We will later also find it convenient
to write the parameters $\theta$ as a $B \times D$ matrix of weights $W$.

@@

To complete the description of our Bayesian framework,
priors on $\theta$ and $\psi$ must also be specified. 
We place a Gaussian prior on $\theta$ with zero mean
and covariance matrix $\sigma^2_\theta I$, so that
each element of $\theta$ has an independent ${\cal N}(0,\sigma_\theta^2)$
prior, with hyperparameter $\sigma_\theta^2$:
%
\begin{equation}
	p(\theta) \sim \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right).
	\label{eqn:theta-prior}
\end{equation}
%
Interestingly, this choice of prior leads to a particularly
simple distribution on the block membership vector $b$ given $X$.
We show in Appendix~\ref{appdx:b|x} that,
after integrating out $\theta$,
$b$ is uniformly distributed given $X$:
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}.
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is an important simplification as evaluating $p(b | X)$ does not require an expensive Monte Carlo integration over $\theta$ nor does it require the exact value of $X$. 
\citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for 
the priors on the additional microcanonical SBM parameters $\psi$, which we adopt. 
The idea is to write the joint distribution on $(b, e, k)$ as a product of 
conditionals, $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. 
In our case, conditioning on $X$ is also necessary, leading to,
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b),
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
where we used the fact $\psi$ and $X$ are conditionally 
independent given $b$.  The exact form of the prior
$p(\psi|b)$ is described in Appendix \ref{appdx:prior}.
All that concerns the main argument is that it has
an easily computable form.

### 2021-05-28 nips/prelim.tex

We first need a model for community-like structure in a network. For this we adopt the widely-used stochastic block model (SBM).
Each node in the graph belongs to a unique community called a block. The probability that two nodes are connected depends only on the block memberships of each. Specifically, we will use the microcanonical variant of the SBM, proposed by \citet{Peixoto-Bayesian-Microcanonical}. To allow for degree-variability between members of the same block, we adopt the degree-corrected formulation (DC-SBM).

@@

\begin{definition}[Microcanonical DC-SBM]
	\label{defn:microcan-dc-sbm}
	Let $N \geq 1$ denote the number of vertices in an undirected graph. The block memberships are encoded by a vector $b \in [B]^N$,
where $B$ is the number of non-empty blocks.
	Let $e=(e_{rs})$ be the $N\times N$ symmetric matrix of edge counts 
between blocks, so that $e_{rs}$ is the number of edges from block $r$ to 
block $s$ -- or twice that number if $r=s$. 
	Let $k =(k_i)$ denote the vector of length $N$ containing the degree
sequence of the graph, with $k_i$ being the degree of vertex $i$.

>>>>

We first need a model for community-like structure in a network. For this we adopt the widely-used stochastic block model (SBM).
Each node in the graph belongs to a unique community called a block. The probability that two nodes are connected depends only on the block memberships of each. Specifically, we will use the microcanonical variant of the SBM, proposed by \citet{Peixoto-Bayesian-Microcanonical}. To allow for degree-variability between members of the same block, we adopt the degree-corrected formulation (DC-SBM), defined in~(\ref{defn:microcan-dc-sbm}).

@@

\begin{definition}[Microcanonical DC-SBM]
	\label{defn:microcan-dc-sbm}
	Let $N \geq 1$ denote the number of vertices in an undirected graph. The block memberships are encoded by a vector $b \in [B]^N$,
where $B$ is the number of non-empty blocks.
	Let $e=(e_{rs})$ be the $N\times N$ symmetric matrix of edge counts 
between blocks, such that $e_{rs}$ is the number of edges from block $r$ to 
block $s$ -- or twice that number if $r=s$. 
	Let $k =(k_i)$ denote the vector of length $N$ containing the degree
sequence of the graph, with $k_i$ being the degree of vertex $i$.
